---
title: "AM Project - Crypto Trading"
author: "Laura Gullicksen, Erich Gozebina, Daria Palitzsch"
date: "23/05/2025"
output: 
  pdf_document:
    number_sections: false
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{titlesec}
  - \titleformat{\section}{\Large\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\normalsize\bfseries}{\thesubsubsection}{1em}{}
  - \usepackage{float}
  - \usepackage{placeins}
  - \FloatBarrier
  - \usepackage{mdframed}
---

```{r setup, include=FALSE}
#setup chunk to load all required packages

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(skimr)
library(dplyr)
library(slider)
library(DescTools)
# library(stargazer)
# library(ggpubr)
# #library(ggalluvial)
library(lubridate)
# library(benford.analysis)
# library(tidyquant)
# library(plotly)
# library(RcppRoll)
# library(scales)
# library(cowplot)
# library(ggforce)
# library(circlize)
# library(zoo)
library(glmnet)
# library(PerformanceAnalytics)
# library(VGAM)
# library(DescTools)
# library(readxl)
# library(broom)
# library(Quandl)
# library(kableExtra)
# library(flextable)
# library(gridExtra)
# library(tinytex)



## setting the correct time zone
Sys.setlocale("LC_TIME", "English") #set language for dates etc.
Sys.setenv(TZ='UCT') #set timezone!


```

```{r, include = FALSE}
#loading the data 
prices <- read.csv("data/pricedata/hourly_prices_0x1f9840a85d5aF5bf1D1762F925BDADdC4201F984.csv")

#change the data type of the timestamp
prices <- prices %>% 
  mutate(hour_timestamp = as.POSIXct(hour_timestamp, format = "%Y-%m-%d %H:%M")) %>%
  # mutate(hour_timestamp = as.Date(ymd_hms(hour_timestamp))) %>% 
  arrange(hour_timestamp) # sort from old to new

```

\begin{mdframed}
\textbf{Requirements}: Implementation of economically senseful trading strategies using machine learning techniques. \end{mdframed}

## 1. Introduction

## 2. Basic model and statistical analysis

In this section we are implementing a first basic model that classifies **Buy**, **Sell** and **Hold**

```{r, include=FALSE}
## Create dependent variable 
#I want to predict the price in the next hour
#For defining Buy, Sell and Hold we use the 25th and 75th percentile

prices <- prices %>% 
  mutate(
    future_return = lead(close_price) / close_price - 1
  )


```

Analyse: muss ich normieren?

```{r, include = FALSE}

# 1. Übersicht: Dichte und Histogramm
ggplot(prices, aes(x = future_return)) +
  geom_histogram(aes(y = ..density..), bins = 100, fill = "skyblue", alpha = 0.6) +
  geom_density(color = "darkred", size = 1) +
  labs(title = "Verteilung von future_return", x = "future_return", y = "Dichte")

# 2. Boxplot: Ausreißer sichtbar machen
ggplot(prices, aes(y = future_return)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  labs(title = "Boxplot von future_return")

# 3. Kennzahlen
summary(prices$future_return)
quantile(prices$future_return, probs = c(0.01, 0.05, 0.95, 0.99), na.rm = TRUE)

```

We have right-skewed data

```{r, include = FALSE}
# 1%- und 99%-Quantile berechnen
q_low <- quantile(prices$future_return, 0.01, na.rm = TRUE)
q_high <- quantile(prices$future_return, 0.99, na.rm = TRUE)

# winsorisieren: alle Werte unter/über der Schwelle auf Grenzwert setzen
prices <- prices %>%
  mutate(future_return_wins = pmin(pmax(future_return, q_low), q_high))
```

```{r, include=FALSE}
#define percentiles: 
qtiles <- quantile(prices$future_return_wins, probs = c(0.25, 0.75), na.rm = TRUE)

prices <- prices %>% 
  mutate(
    signal = case_when(
      future_return_wins < qtiles[1] ~ "Sell",
      future_return_wins > qtiles[2] ~ "Buy",
      TRUE ~ "Hold"
    )
  )


#TODO: Alternativ rollierendes quantil (z.B. über 90 Stunden)

```

```{r, include=FALSE}
#Define Features: 
prices <- prices %>% 
  mutate(
      mean_24h = close_price / slide_dbl(close_price, mean, .before=23, .complete = TRUE),
      sd_24h = slide_dbl(close_price, sd, .before=23, .complete = TRUE),
      return_24h = close_price / lag(close_price, 24) -1,
      fluctuation = high_price - low_price
  )


```

```{r, include = FALSE}
# prices <- prices %>% 
#   mutate(
#     fluctuation = high_price - low_price, 
#     difference = close_price - open_price, 
#     return = close_price / lag(close_price) - 1,
#     log_return = log(return), 
#     #intraday_return = difference / open_price, 
#     hour = lubridate::hour(hour_timestamp),
#     
#     pred_next_closeprice = lead(close_price) - close_price,
#     
#     pred_sign_next_closeprice = case_when( 
#       pred_next_closeprice < 0 ~ -1, #goes down
#       pred_next_closeprice > 0 ~ 1, # goes up
#       is.na(pred_next_closeprice) ~ NA,
#       TRUE ~ 0)#stays the same
#     
#     #normierter next_return
# 
#   )


#add senseful columns -> e.g. day difference (relative to daily open_price)

```

**Outlier händeln**

```{r, include=FALSE}

# Ziel: signal (z. B. "Buy", "Sell", "Hold")
# Features: fluctuation, close_price, mean_24h
# Daten vorbereiten
prices <- na.omit(prices)

# Designmatrix & Zielvariable
x <- model.matrix(signal ~ fluctuation + close_price + mean_24h + sd_24h, prices)[, -1]  # Ohne Intercept
y <- prices$signal

# Train/Test-Split (50/50)
split_index <- floor(nrow(prices) * 0.8)
train.rows <- 1:split_index
test.rows  <- (split_index + 1):nrow(prices)

# Lasso-Grid für Lambda
grid <- 10^seq(5, -5, length = 100)

# Lasso-Modell trainieren
set.seed(1)
lasso.mod <- glmnet(x[train.rows, ], y[train.rows], 
                    family = "multinomial", alpha = 1, lambda = grid)
plot(lasso.mod, xvar = "lambda")

# Cross-Validation zur Auswahl von lambda.min
cv.out <- cv.glmnet(x[train.rows, ], y[train.rows], 
                    alpha = 1, family = "multinomial")
plot(cv.out)
bestlam <- cv.out$lambda.min
cat("Bestes Lambda:", bestlam, "\n")

# Prognose auf Testdaten (Wahrscheinlichkeiten)
lasso.pred <- predict(lasso.mod, s = bestlam, newx = x[test.rows, ], type = "response")

# Wahrscheinlichkeiten → Klassenvorhersage
# Achtung: lasso.pred ist ein Array mit 3 Dimensionen
lasso.class <- apply(lasso.pred[,,1], 1, function(row) colnames(lasso.pred)[which.max(row)])

# Konfusionsmatrix
confusion <- table(Predicted = lasso.class, Actual = y[test.rows])
print(confusion)

# Optional: Genauigkeit berechnen
accuracy <- mean(lasso.class == y[test.rows])
cat("Genauigkeit:", round(accuracy, 4), "\n")

```

```{r, include = FALSE}
#
prices = na.omit(prices)
#x = model.matrix(pred_sign_next_closeprice ~ fluctuation + close_price, prices)[,-1]
x = model.matrix(signal ~ fluctuation + close_price + mean_24h, prices)[,-1]
y = prices$signal

train.rows = 1:(nrow(prices)/2) #momentan 50:50 -> anpassen 
grid=10^seq(5,-5,length=100)
set.seed(1)

lasso.mod = glmnet(x[train.rows, ], y[train.rows], family = "multinomial", alpha = 1, lambda = grid) 
plot(lasso.mod, xvar = "lambda")


 
#cross validation (10-fold)
cv.out=cv.glmnet(x[train.rows,],y[train.rows],alpha=1, family = "multinomial") #alpha = 1 -> Lasso
plot(cv.out)
bestlam=cv.out$lambda.min
bestlam
log(bestlam)

test.rows = round((nrow(prices)/2 + 1),0):(nrow(prices)-1)

lasso.pred=predict(lasso.mod,s=bestlam,newx=x[test.rows,])
#mean((lasso.pred-y[test.rows])^2)
pred_table <- table(Predicted = lasso.pred, Actual = y[1:30920]) %>% 
  as.data.frame()

# full sample 
# out=glmnet(x,y,alpha=1,lambda=grid, family = "multinomial") 
# lasso.coef=predict(out,type="coefficients",s=bestlam)[1:20,]
# lasso.coef
# lasso.coef[lasso.coef!=0]

```

```{r}

```

### 2.1 Approach

### 2.2 Implementation

**Backtesting here? or in another section**

**Backtesting Strategies:** -\>Buy&Hold etc

### 2.3 Results

### 2.4 Interpretation

## 3. Extensions

### 3.1 Approaches

### 3.2 Implementation

**Backtesting here? or in another section**

### 3.3 Results

### 3.4 Interpretation

## 4. Conclusion
