---
title: "AM Project - Crypto Trading"
author: "Laura Gullicksen, Erich Gozebina, Daria Palitzsch"
date: "23/05/2025"
output: 
  pdf_document:
    number_sections: false
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{titlesec}
  - \titleformat{\section}{\Large\bfseries}{\thesection}{1em}{}
  - \titleformat{\subsection}{\large\bfseries}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\normalsize\bfseries}{\thesubsubsection}{1em}{}
  - \usepackage{float}
  - \usepackage{placeins}
  - \FloatBarrier
  - \usepackage{mdframed}
---

```{r setup, include=FALSE}
#setup chunk to load all required packages

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ggplot2)
library(skimr)
library(dplyr)
# library(stargazer)
# library(ggpubr)
# #library(ggalluvial)
# library(lubridate)
# library(benford.analysis)
# library(tidyquant)
# library(plotly)
# library(RcppRoll)
# library(scales)
# library(cowplot)
# library(ggforce)
# library(circlize)
# library(zoo)
# library(glmnet)
# library(PerformanceAnalytics)
# library(VGAM)
# library(DescTools)
# library(readxl)
# library(broom)
# library(Quandl)
# library(kableExtra)
# library(flextable)
# library(gridExtra)
# library(tinytex)



## setting the correct time zone
Sys.setlocale("LC_TIME", "English") #set language for dates etc.
Sys.setenv(TZ='UCT') #set timezone!


```


```{r, include = FALSE}
#loading the data 
prices <- read.csv("data/pricedata/hourly_prices_bitcoin.csv")

#change the data type of the timestamp
prices <- prices %>% 
  mutate(hour_timestamp = as.POSIXct(hour_timestamp, format = "%Y-%m-%d %H:%M")) %>%
  arrange(hour_timestamp) # sort from old to new
```


\begin{mdframed}
\textbf{Requirements}: Implementation of economically senseful trading strategies using machine learning techniques. \end{mdframed}

## 1. Introduction

## 2. Basic model and statistical analysis

### 2.1 Approach

### 2.2 Implementation

**Backtesting here? or in another section**

### 2.3 Results 

### 2.4 Interpretation

## 3. Extensions

### 3.1 Approaches

### 3.2 Implementation

**Backtesting here? or in another section**

### 3.3 Results 

### 3.4 Interpretation 

## 4. Conclusion
